---
title: "Gene Expression Analysis: Preprocessing, Clustering, Classification & Model Evaluation"
author: "Promise"
date: "2026-08-12"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Step 1: Load libraries
```{r}
library(readr)
```

Step 2: Load and reshape data
```{r}
brca <- read.csv("data/brca.csv", header = FALSE, stringsAsFactors = FALSE)
dim(brca)

sample_ids <- brca[1, -1]
measure_type <- brca[2, -1]
gene_names <- brca[-c(1,2), 1]

keep_cols <- which(measure_type == "RPKM") + 1

counts <- brca[-c(1,2), keep_cols]
counts <- apply(counts, 2, as.numeric)
rownames(counts) <- gene_names
colnames(counts) <- as.character(sample_ids[measure_type == "RPKM"])

dim(counts)
counts[1:5, 1:5]

brca <- counts
rm(counts)
dim(brca)
```

Step 3: Clean the data
```{r}
keep <- rowSums(brca > 0) / ncol(brca) >= 0.2
brca <- brca[keep, ]

brca <- brca[apply(brca, 1, var) > 0, ]

sum(is.na(brca))
brca[is.na(brca)] <- 0

dim(brca)
```

Step 4: Descriptive statistics
```{r}
gene_mean   <- rowMeans(brca)
gene_median <- apply(brca, 1, median)
gene_sd     <- apply(brca, 1, sd)

summary(gene_mean)
summary(gene_median)
summary(gene_sd)

sample_totals <- colSums(brca)
summary(sample_totals)
```

Step 5: Log2 transform
```{r}
log_brca <- log2(brca + 1)
```

Step 6: Visualize
```{r}
boxplot(log_brca[, 1:20],
        col = colorRampPalette(c("#a1dab4", "#41b6c4", "#2c7fb8"))(20),
        border = "gray30",
        las = 2,
        cex.axis = 0.7,
        names = substr(colnames(log_brca)[1:20], 1, 12),
        main = "Expression per Sample (log2 RPKM)",
        ylab = "log2(RPKM + 1)",
        outline = TRUE,
        outcol = "gray50",
        outpch = 21,
        outbg = "gray80",
        outcex = 0.6,
        boxwex = 0.6,
        whisklty = 1,
        staplewex = 0.3)

grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")
box()

hist(log_brca, breaks = 50, col = "steelblue",
     main = "Histogram of log2(RPKM + 1)",
     xlab = "log2(RPKM + 1)")

gene_var <- apply(log_brca, 1, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:50]
heatmap(log_brca[top_genes, 1:20])
```

Step 7: PCA
```{r}
pca_data_full <- t(log_brca)
pca_result_full <- prcomp(pca_data_full, scale. = TRUE)

var_explained_full <- (pca_result_full$sdev^2) / sum(pca_result_full$sdev^2) * 100

barplot(var_explained_full[1:10],
        names.arg = paste0("PC", 1:10),
        col = colorRampPalette(c("#2c7fb8", "#41b6c4", "#a1dab4"))(10),
        main = "Scree Plot: % Variance Explained (All Genes)",
        ylab = "Percentage of Variance Explained (%)",
        xlab = "Principal Component",
        border = "white", las = 1)
text(x = seq(0.7, by = 1.2, length.out = 10),
     y = var_explained_full[1:10] + 0.5,
     labels = paste0(round(var_explained_full[1:10], 1), "%"),
     cex = 0.8)

sample_type <- substr(colnames(brca), 14, 15)
group <- ifelse(sample_type == "01", "Tumor", "Normal")
group_colors <- ifelse(group == "Tumor", "#e34a33", "#2c7fb8")

plot(pca_result_full$x[,1], pca_result_full$x[,2],
     col = group_colors, pch = 19, cex = 1.2,
     xlab = paste0("PC1 (", round(var_explained_full[1], 1), "% variance)"),
     ylab = paste0("PC2 (", round(var_explained_full[2], 1), "% variance)"),
     main = "PCA Plot: Tumor vs Normal Samples (All Genes)",
     bty = "l")
grid(col = "gray90")
abline(h = 0, v = 0, col = "gray70", lty = 2)
legend("topright", legend = c("Tumor", "Normal"),
       col = c("#e34a33", "#2c7fb8"), pch = 19, bty = "n")
```

Step 8: Load additional libraries
```{r}
library(cluster)
library(caret)
library(randomForest)
library(e1071)
library(class)
library(pROC)
library(glmnet)
```

Step 9: K-means clustering
```{r}
var_explained_full <- (pca_result_full$sdev^2) / sum(pca_result_full$sdev^2) * 100
cum_var <- cumsum(var_explained_full)
n_pcs <- which(cum_var >= 80)[1]
cat("Using", n_pcs, "PCs, explaining", round(cum_var[n_pcs], 1), "% of variance\n")

km_data_sc <- pca_result_full$x[, 1:n_pcs]

set.seed(123)
wss <- sapply(1:8, function(k) kmeans(km_data_sc, centers = k, nstart = 10)$tot.withinss)
plot(1:8, wss, type = "b", pch = 19,
     xlab = "Number of clusters (k)", ylab = "Total within-cluster SS",
     main = "Elbow Method for Choosing k")

set.seed(123)
km_res <- kmeans(km_data_sc, centers = 2, nstart = 25)

table(Cluster = km_res$cluster, TrueLabel = group)

plot(km_data_sc[, 1], km_data_sc[, 2],
     col = c("#2c7fb8", "#e34a33")[km_res$cluster],
     pch = 19, xlab = "PC1", ylab = "PC2",
     main = paste0("K-means Clustering (k = 2) on Top ", n_pcs, " PCs"))
legend("topright", legend = paste("Cluster", sort(unique(km_res$cluster))),
       col = c("#2c7fb8", "#e34a33"), pch = 19, bty = "n")
```

Step 10: Hierarchical clustering
```{r}
d <- dist(km_data_sc, method = "euclidean")
hc <- hclust(d, method = "ward.D2")

plot(hc, labels = FALSE, main = "Hierarchical Clustering Dendrogram",
     xlab = "Samples", sub = "")
hc_clusters <- cutree(hc, k = 2)
rect.hclust(hc, k = 2, border = c("#2c7fb8", "#e34a33"))

table(Cluster = hc_clusters, TrueLabel = group)
```

Step 11: Silhouette score
```{r}
sil <- silhouette(km_res$cluster, d)
plot(sil, col = c("#2c7fb8", "#e34a33"), main = "Silhouette Plot (K-means, k = 2)")
cat("Average silhouette width (k-means):", mean(sil[, 3]), "\n")
```

Step 12: Consensus clustering - feature selection
```{r}
gene_var_cc <- apply(log_brca, 1, var)
top_var_cc  <- names(sort(gene_var_cc, decreasing = TRUE))[1:1000]

cc_data    <- t(log_brca[top_var_cc, ])
cc_data_sc <- scale(cc_data)
cc_data_sc <- cc_data_sc[, colSums(is.na(cc_data_sc)) == 0]
dim(cc_data_sc)
```

Step 13: Consensus clustering - k-means base
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("ConsensusClusterPlus", quietly = TRUE)) {
  BiocManager::install("ConsensusClusterPlus", update = FALSE, ask = FALSE)
}
if (!requireNamespace("mclust", quietly = TRUE)) {
  install.packages("mclust")
}

library(ConsensusClusterPlus)

consensus_input <- t(cc_data_sc)

results_cc <- ConsensusClusterPlus(
  consensus_input,
  maxK       = 6,
  reps       = 100,
  pItem      = 0.8,
  pFeature   = 1,
  clusterAlg = "km",
  distance   = "euclidean",
  innerLinkage = "ward.D2",
  finalLinkage  = "ward.D2",
  seed       = 123,
  plot       = NULL,
  title      = "consensus_cluster_output"
)
```

Step 14: Consensus clustering - hierarchical base
```{r}
results_cc_hc <- ConsensusClusterPlus(
  consensus_input,
  maxK       = 6,
  reps       = 100,
  pItem      = 0.8,
  pFeature   = 1,
  clusterAlg = "hc",
  distance   = "euclidean",
  innerLinkage = "ward.D2",
  finalLinkage  = "ward.D2",
  seed       = 123,
  plot       = NULL,
  title      = "consensus_cluster_output_hc"
)
```

Step 15: Choosing k (PAC)
```{r}
calc_pac <- function(consensus_matrix, lower = 0.1, upper = 0.9) {
  vals <- consensus_matrix[upper.tri(consensus_matrix)]
  mean(vals > lower & vals < upper)
}

pac_vals <- sapply(2:6, function(k) calc_pac(results_cc[[k]]$consensusMatrix))
names(pac_vals) <- paste0("k=", 2:6)
pac_vals

best_k <- as.integer(sub("k=", "", names(which.min(pac_vals))))
cat("Selected k (lowest PAC):", best_k, "\n")

plot(2:6, pac_vals, type = "b", pch = 19, xlab = "k",
     ylab = "Proportion of Ambiguous Clustering (PAC)",
     main = "PAC by Number of Clusters")
abline(v = best_k, col = "red", lty = 2)
```

Step 16: Final consensus clusters and agreement with true labels
```{r}
consensus_clusters <- results_cc[[best_k]]$consensusClass
names(consensus_clusters) <- colnames(consensus_input)

table(ConsensusCluster = consensus_clusters, TrueLabel = group)

mean_within_between <- function(consensus_matrix, clusters) {
  n <- length(clusters)
  within_vals <- c(); between_vals <- c()
  for (i in 1:(n - 1)) for (j in (i + 1):n) {
    v <- consensus_matrix[i, j]
    if (clusters[i] == clusters[j]) within_vals <- c(within_vals, v)
    else between_vals <- c(between_vals, v)
  }
  c(mean_within = mean(within_vals), mean_between = mean(between_vals))
}
mean_within_between(results_cc[[best_k]]$consensusMatrix, consensus_clusters)
```

Step 17: Consensus clustering - ARI
```{r}
library(mclust)

ari_vs_truth <- adjustedRandIndex(consensus_clusters, group)
ari_km_vs_hc <- adjustedRandIndex(
  results_cc[[best_k]]$consensusClass,
  results_cc_hc[[best_k]]$consensusClass
)

cat("ARI: consensus clusters vs true Tumor/Normal labels:", round(ari_vs_truth, 4), "\n")
cat("ARI: k-means-based vs hierarchical-based consensus solution:", round(ari_km_vs_hc, 4), "\n")
```

Step 18: Feature selection for classification
```{r}
group_factor <- factor(group, levels = c("Normal", "Tumor"))

t_stats <- apply(log_brca, 1, function(x) {
  tt <- try(t.test(x ~ group_factor), silent = TRUE)
  if (inherits(tt, "try-error")) return(0)
  tt$statistic
})
top_features <- names(sort(abs(t_stats), decreasing = TRUE))[1:100]

model_data <- as.data.frame(t(log_brca[top_features, ]))
model_data$Class <- group_factor

colnames(model_data) <- make.names(colnames(model_data), unique = TRUE)

dim(model_data)
table(model_data$Class)
```

Step 19: Train/test split
```{r}
set.seed(42)
train_idx  <- createDataPartition(model_data$Class, p = 0.7, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

dim(train_data); dim(test_data)
table(train_data$Class)
table(test_data$Class)
```

Step 20: Train Random Forest
```{r}
train_x_rf <- train_data[, setdiff(names(train_data), "Class")]
train_y_rf <- train_data$Class

set.seed(42)
rf_model <- randomForest(x = train_x_rf, y = train_y_rf, ntree = 500,
                          importance = TRUE)
print(rf_model)

varImpPlot(rf_model, n.var = 15, main = "Top 15 Genes by Importance (RF)")
```

Step 21: Train SVM
```{r}
set.seed(42)
svm_model <- svm(Class ~ ., data = train_data, kernel = "radial",
                  probability = TRUE)
summary(svm_model)
```

Step 22: Train kNN
```{r}
train_x <- train_data[, setdiff(names(train_data), "Class")]
test_x  <- test_data[,  setdiff(names(test_data),  "Class")]

set.seed(42)
knn_pred <- knn(train = train_x, test = test_x,
                 cl = train_data$Class, k = 5, prob = TRUE)
```

Step 23: Train Elastic Net Logistic Regression
```{r}
x_train <- as.matrix(train_data[, setdiff(names(train_data), "Class")])
y_train <- train_data$Class
x_test  <- as.matrix(test_data[,  setdiff(names(test_data),  "Class")])

set.seed(42)
cv_fit <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 0.5)
plot(cv_fit)
```

Step 24: Predictions on test set
```{r}
rf_pred   <- predict(rf_model, newdata = test_data)
svm_pred  <- predict(svm_model, newdata = test_data)
glm_prob  <- predict(cv_fit, newx = x_test, s = "lambda.min", type = "response")
glm_pred  <- factor(ifelse(glm_prob > 0.5, "Tumor", "Normal"),
                     levels = levels(test_data$Class))
```

Step 25: Confusion matrices
```{r}
cm_rf   <- confusionMatrix(rf_pred,  test_data$Class, positive = "Tumor")
cm_svm  <- confusionMatrix(svm_pred, test_data$Class, positive = "Tumor")
cm_knn  <- confusionMatrix(knn_pred, test_data$Class, positive = "Tumor")
cm_glm  <- confusionMatrix(glm_pred, test_data$Class, positive = "Tumor")

cm_rf
cm_svm
cm_knn
cm_glm
```

Step 26: Accuracy comparison table
```{r}
extract_metrics <- function(cm, model_name) {
  data.frame(
    Model       = model_name,
    Accuracy    = round(cm$overall["Accuracy"], 4),
    Kappa       = round(cm$overall["Kappa"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    Precision   = round(cm$byClass["Precision"], 4),
    F1          = round(cm$byClass["F1"], 4)
  )
}

results <- rbind(
  extract_metrics(cm_rf,  "Random Forest"),
  extract_metrics(cm_svm, "SVM (RBF)"),
  extract_metrics(cm_knn, "kNN (k=5)"),
  extract_metrics(cm_glm, "Elastic Net Logistic")
)
rownames(results) <- NULL
results
```

Step 27: Accuracy comparison plot
```{r}
barplot(results$Accuracy, names.arg = results$Model,
        col = c("#2c7fb8", "#41b6c4", "#a1dab4", "#e34a33"),
        ylim = c(0, 1), las = 2, ylab = "Test-set Accuracy",
        main = "Model Accuracy Comparison")
abline(h = seq(0, 1, 0.2), col = "gray90", lty = "dotted")
box()
```

Step 28: ROC curves and AUC
```{r}
rf_prob  <- predict(rf_model, newdata = test_data, type = "prob")[, "Tumor"]
svm_score <- attr(predict(svm_model, newdata = test_data, probability = TRUE),
                   "probabilities")[, "Tumor"]
knn_score <- ifelse(knn_pred == "Tumor",
                     attr(knn_pred, "prob"), 1 - attr(knn_pred, "prob"))

roc_rf  <- roc(test_data$Class, rf_prob,  levels = c("Normal", "Tumor"))
roc_svm <- roc(test_data$Class, svm_score, levels = c("Normal", "Tumor"))
roc_knn <- roc(test_data$Class, knn_score, levels = c("Normal", "Tumor"))
roc_glm <- roc(test_data$Class, as.numeric(glm_prob), levels = c("Normal", "Tumor"))

plot(roc_rf,  col = "#2c7fb8", lwd = 2, main = "ROC Curves: Model Comparison")
lines(roc_svm, col = "#41b6c4", lwd = 2)
lines(roc_knn, col = "#a1dab4", lwd = 2)
lines(roc_glm, col = "#e34a33", lwd = 2)
legend("bottomright",
       legend = c(paste0("RF (AUC=",  round(auc(roc_rf), 3), ")"),
                  paste0("SVM (AUC=", round(auc(roc_svm), 3), ")"),
                  paste0("kNN (AUC=", round(auc(roc_knn), 3), ")"),
                  paste0("GLMNET (AUC=", round(auc(roc_glm), 3), ")")),
       col = c("#2c7fb8", "#41b6c4", "#a1dab4", "#e34a33"), lwd = 2, bty = "n")
```

Step 29: 5-fold cross-validation (Random Forest)
```{r}
ctrl <- trainControl(method = "cv", number = 5, classProbs = TRUE,
                      summaryFunction = twoClassSummary, savePredictions = TRUE)

set.seed(42)
rf_cv <- train(Class ~ ., data = train_data, method = "rf",
                trControl = ctrl, metric = "ROC",
                tuneGrid = expand.grid(mtry = c(2, 5, 10, 15)))

rf_cv
plot(rf_cv, main = "5-fold CV: Accuracy (ROC) vs mtry")
```

Step 30: Summary
```{r}
cat("Clustering agreement (k-means vs true labels):\n")
print(table(km_res$cluster, group))

cat("\nBest performing model by test-set accuracy:\n")
print(results[which.max(results$Accuracy), ])

cat("\nBest cross-validated mtry for Random Forest:", rf_cv$bestTune$mtry, "\n")
```
